---
title: 'Reproducible Research: Peer Assessment 1'
author: "Silverio Neves"
date: "24 September 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setoptions, echo=TRUE}
```


## Loading and preprocessing the data
Load data from "activity.csv" in "activity" data.frame
```{r data}
if(!require(readr)) install.packages("readr")
library(readr)
activity <- read_csv("activity.csv")
d <- dim(activity)
```
Data.frame with `r d[1]` observations and `r d[2]` variables ("steps", "date","interval")

## What is mean total number of steps taken per day?

```{r numberstepsday}
if(!require(dplyr)) install.packages("dplyr")
library(dplyr)
activityday <- group_by(activity, date)
sumstepsday <- dplyr::summarise(activityday, stepsday = sum(steps, na.rm = TRUE))
m <- as.integer(mean(sumstepsday$stepsday, na.rm = TRUE))
t <- as.integer(sum(sumstepsday$stepsday, na.rm = TRUE))
med <- as.integer(median(sumstepsday$stepsday, na.rm = TRUE))
```

In the figure we can see the histogram of the total number of steps taken each day

```{r histogrambase}
if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)
qplot(sumstepsday$stepsday, geom="histogram")+
        labs(title = "Histogram steps by day", subtitle = "days with no data (NA) are excluded",
                y = "Number of occurences", caption = "observations from 2012-10-01 to 2012-11-30",
                x = "Total steps per day")
```

In **average** they do **`r m`** steps per day, with a **median** of **`r med`** for a **total** of **`r t`** steps, excluding days without steps observation (= NA)

We can see in the plot bellow the behaviour (steps by day). A time series plot of the average number of steps taken.

```{r graphicstepsbyday}
if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)
ggplot(data=sumstepsday,
        aes(x=date, y=stepsday)) +
        geom_line() +
        geom_point() +
        scale_color_brewer(palette="Paired")+
        theme_minimal()+
        theme(legend.position="right")+
        labs(title = "Total steps by day", subtitle = "days with no data (NA) are excluded",
                x = "day", caption = "observations from 2012-10-01 to 2012-11-30",
                y = "Total steps")

```


## What is the average daily activity pattern?

```{r numberstepsinterval}
if(!require(dplyr)) install.packages("dplyr")
library(dplyr)
activityinterval <- group_by(activity, interval)
sumstepsinterval <- dplyr::summarise(activityinterval, stepsinterval = mean(steps, na.rm = TRUE))
maxiinterval <- order(sumstepsinterval$stepsinterval, decreasing = TRUE)
```
The **interval** with the **maximum steps in average** is **`r as.integer(sumstepsinterval[maxiinterval[1],1])`** with an **average** of **`r as.integer(sumstepsinterval[maxiinterval[1],2])`**.

```{r graphicstepsbyinterval}
if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)
ggplot(data=sumstepsinterval,
        aes(x=interval, y=stepsinterval)) +
        geom_line() +
        geom_point() +
        scale_color_brewer(palette="Paired")+
        theme_minimal()+
        theme(legend.position="right")+
        labs(title = "Total steps by interval", subtitle = "days with no data (NA) are excluded",
                x = "interval", caption = "observations from 2012-10-01 to 2012-11-30",
                y = "Total steps")

```


## Imputing missing values

```{r number.na}
totalrows <- dim(activity)[1]
c <- sum(complete.cases(activity))
s <- sum(is.na(activity$steps))
```
The **database** has **`r c` complete cases** (rows without missing values). From the "is.na" function we can see that **variable "steps"** has **`r s` NA's values** which are precise the difference between total rows of `r totalrows' and complete cases.
  
We will **fill NAs values** with the **mean** for the specific **interval** creating a new data.frame "newactivity"

```{r fillna}
newactivity <- activity
ridna <- which(is.na(newactivity$steps))
for(i in ridna){
        newactivity$steps[i] <- sumstepsinterval$stepsinterval[newactivity$interval[i] == sumstepsinterval$interval] 
}
```

```{r newnumberstepsday}
if(!require(dplyr)) install.packages("dplyr")
library(dplyr)
newactivityday <- group_by(newactivity, date)
newsumstepsday <- dplyr::summarise(newactivityday, stepsday = sum(steps, na.rm = TRUE))
newm <- as.integer(mean(newsumstepsday$stepsday, na.rm = TRUE))
newt <- as.integer(sum(newsumstepsday$stepsday, na.rm = TRUE))
newmed <- as.integer(median(newsumstepsday$stepsday, na.rm = TRUE))
```
In **average** they do **`r newm`** steps per day, with a **median** of **`r newmed`** for a **total** of **`r newt`** steps, excluding days without steps observation (= NA)

All values grew. Mean from `r m` to `r newm`, and the median from `r med` to `r newmed`. For curiosity, the new mean is equal to the new median.

In the figure we can see the new histogram of the total number of steps taken each day

```{r histogramnew}
if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)
qplot(newsumstepsday$stepsday, geom="histogram")+
        labs(title = "Histogram steps by day", subtitle = "days with no data (NA) were included by replacing value with interval mean",
                y = "Number of occurences", caption = "observations from 2012-10-01 to 2012-11-30",
                x = "Total steps per day")
```


We can see in the plot bellow the behaviour (steps by day)

```{r newgraphicstepsbyday}
if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)
ggplot(data=newsumstepsday,
        aes(x=date, y=stepsday)) +
        geom_line() +
        geom_point() +
        scale_color_brewer(palette="Paired")+
        theme_minimal()+
        theme(legend.position="right")+
        labs(title = "Total steps by day", subtitle = "days with no data (NA) were included by replacing value with interval mean",
                x = "day", caption = "observations from 2012-10-01 to 2012-11-30",
                y = "Total steps")

```


## Are there differences in activity patterns between weekdays and weekends?

First it is necessary to create a new collumn to distinguish from weekday and weekend day. New column has the name "day"
```{r numberstepsweek}
if(!require(dplyr)) install.packages("dplyr")
library(dplyr)

days <- c("sábado", "domingo") ## need to change for local and system language day names
newactivity$day <- 0
for (i in 1:dim(newactivity)[1])
{
        if (weekdays(newactivity$date[i]) %in% days) {newactivity$day[i] = "weekend"} else {newactivity$day[i] = "weekday"}
}
```

Now we will generate a graphic to compare weekday and weekend days performance
```{r numberstepsweekreport}
if(!require(dplyr)) install.packages("dplyr")
library(dplyr)
newactivityweek <- group_by(newactivity, interval, day)
newsumstepsweek <- dplyr::summarise(newactivityweek, stepsinterval = sum(steps, na.rm = TRUE))

if(!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)
ggplot(data=newsumstepsweek,
        aes(x=interval, y=stepsinterval, col = day)) +
        geom_line() +
        geom_point() +
        facet_grid(day~.) +
        scale_color_brewer(palette="Paired")+
        theme_minimal()+
        theme(legend.position="right")+
        labs(title = "Total steps by interval", subtitle = "days with no data (NA) were included by replacing value with interval mean",
                x = "interval", caption = "observations from 2012-10-01 to 2012-11-30",
                y = "Total steps")
```

